---
author: "Sean Anderson, Paul Regular"
title: "Example of simulating data with SimSurvey and index standardization with sdmTMB"
---

# Setup

```{r, warning=FALSE, message=FALSE}
library(raster)
library(dplyr)
library(ggplot2)
library(sdmTMB)
library(SimSurvey)
library(tidyverse)

theme_set(gfplot::theme_pbs())
```

# Population and survey simulation

First we will simulate a spatial, age-structured population with SimSurvey. 

```{r}
set.seed(32)
 sim <- sim_abundance(ages = seq(1, 10), years = seq(1, 10)) %>%
    sim_distribution(
      grid = make_grid(res = c(10, 10),
        # method == "bezier",
        # x_range = c(-240, 240), y_range = c(-240, 240), # to make bigger grid
        # shelf_width = 100, # to make narrower shelf
        strat_splits = 1,
        strat_breaks = seq(0, 1000, by = 100), #to make fewer strat for design-based
        depth_range = c(10, 400)), #10, 500 was example
      ays_covar = sim_ays_covar(
        sd = 2, #2.8 is default
        #lambda = 1, # 1 is default, changes degree of smoothness
        range = 80, # make less spatial correlation, from default of 300
        phi_age = 0.8, phi_year = 0.1),
      depth_par = sim_parabola(mu = 200, sigma = 5, plot = F) # simga = 30 was example
    )
```

Simulated depth profile of a wide flat shelf
```{r}
  ## View(sim$grid_xy)
  sim$grid_xy %>%
    ggplot(aes(x, y, fill = depth)) +
    geom_raster() +
    scale_fill_viridis_c(
      limits=c(150, 250),
      direction = -1)
```

# make each strat for design-based span 100m
```{r}
  sim$grid_xy %>%
    ggplot(aes(x, y, fill = as.factor(strat))) +
    geom_raster() +
    scale_fill_viridis_d(
      #limits=c(150, 250),
      direction = -1)
```

Include a narrower depth preference and reduced range parameter for spatial field in order to simulate a distribution that shows a pattern in response to depth not just at the edges
```{r}
sim$sp_N %>% left_join(sim$grid_xy) %>%
    group_by(x, y, year, cell) %>%
    summarise(N = sum(N), .groups = "drop") %>%
    ggplot(aes(x, y,
      fill = log(N + 1)
      )) +
    geom_raster() +
    facet_wrap(~year) +
    scale_fill_viridis_c()
```

Then we simulate survey samples from this population

```{r}
  survey <- sim_survey(sim, n_sims = 1, #min_sets = 10
    ) %>% run_strat()
  xy <- as_tibble(survey$grid_xy)
  dat <- as_tibble(survey$setdet) %>%
    dplyr::select(x, y, set, year, N = n, tow_area)
  dat <- left_join(dat, xy, by = c("x", "y"))
  dat$offset <- log(dat$tow_area)

  grid_dat <- as_tibble(dplyr::select(sim$grid_xy, x, y, depth)) %>% distinct()
  grid_dat <- purrr::map_dfr(sort(unique(dat$year)), ~ bind_cols(grid_dat, year = .))
  grid_dat$offset <- mean(dat$offset)
```

Here are the survey observations:

```{r}
ggplot(dat, aes(x, y, colour = log(N + 1), size = N)) +
  geom_point() +
  facet_wrap(vars(year)) +
  scale_colour_viridis_c() +
  scale_size_area(max_size = 4)
```



# Fitting a geostatistical spatiotemporal model


```{r}
mesh <- sdmTMB::make_mesh(dat, xy_cols = c("x", "y"), cutoff = 20)
```

We can make a plot of our mesh:

```{r}
plot(mesh)
```

Fit model with a fixed spatial field that might account for depth even without a true depth covariate
```{r fit, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
  fit <- sdmTMB(N ~ 0 + as.factor(year) + offset,
    data = dat,
    family = nbinom2(link = "log"), spde = mesh,
    include_spatial = TRUE,
    time = "year"
  )
```

```{r}
print(fit)
```

Fit model with depth, but without a fixed spatial field
```{r fit-depth, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
 fit_depth <- sdmTMB(N ~ 0 + as.factor(year) + s(depth, k = 3) +
      offset,
    data = dat,
    family = nbinom2(link = "log"), spde = mesh,
    # include_spatial = TRUE,
    time = "year"
  )
```

```{r}
print(fit_depth)
```


```{r}
nd <- data.frame(depth = seq(min(dat$depth), max(dat$depth), length.out = 80))
  # add all years in case of time-varying depth
  nd <- purrr::map_dfr(sort(unique(dat$year)), ~ bind_cols(nd, year = .))
  nd$offset <- mean(dat$offset)
  
  p1 <- predict(fit_depth,
    newdata = filter(nd, year == nd$year[1]),
    se_fit = TRUE, re_form = NA)
  
 y <- sim_parabola(mu = 200, sigma = 5, plot=F)(x = 50:360)
  tru_curve <- data.frame(x = 50:360, y)  

  ggplot(p1, aes(depth, est,
    ymin = est - 1.96 * est_se, ymax = est + 1.96 * est_se)) +
    geom_line() + geom_ribbon(alpha = 0.4) +
    geom_line(data = tru_curve, aes(y=y, x = x),
      colour = "red", linetype = "dashed",
      inherit.aes = F)
```


And extract values from estimated fixed and random effect variables using the `tidy()` method, which follows the broom package convention:

```{r}
tidy(fit_depth, conf.int = TRUE)
tidy(fit_depth, conf.int = TRUE, effects = "ran_pars")
```

We can look at residuals:

```{r}
dat$resid <- residuals(fit_depth)

ggplot(dat, aes(x, y, colour = resid, size = abs(resid))) +
  geom_point() +
  facet_wrap(~year) +
  scale_colour_gradient2() +
  scale_size_area(max_size = 2)
```

These are randomized quantile residuals---similar to the DHARMa R package. These are adjusted so that they would be Gaussian distributed if the model were consistent with the data:

```{r}
hist(dat$resid)
```

# Predicting over the survey domain and calculating an index of abundance

To calculate an index of abundance, first we need to create a grid that covers the entirety of our survey domain. In this case that would be a square grid:

```{r}
pred <- predict(fit, newdata = grid_dat, return_tmb_object = TRUE, area = 100)
pred_depth <- predict(fit_depth, newdata = grid_dat, return_tmb_object = TRUE, area = 100)
index <- get_index(pred)
index_depth <- get_index(pred_depth)

true_abund <- tibble(year = unique(dat$year), N = as.numeric(colSums(survey$I))) %>%
    mutate(type = "True")

strat_abund <- tibble::as_tibble(survey$total_strat) %>%
    mutate(N = total, type = "Design-based") %>%
    dplyr::select(year, N, type)

result <- mutate(index, type = "Model-based", N = est) %>%
    bind_rows(mutate(index_depth, type = "Model-based-depth", N = est)) %>%
    bind_rows(strat_abund) %>%
    bind_rows(true_abund) 

result_scaled <- result %>%
  group_by(type) %>%
  mutate(geo_mean = exp(mean(log(N), na.rm = TRUE)),
    lwr_scaled = lwr / geo_mean, N_scaled = N / geo_mean, upr_scaled = upr / geo_mean,
    type = factor(type, levels = c("True", "Design-based", "Model-based", "Model-based-depth"))) %>%
  ungroup() %>%
  arrange(type)

result_scaled %>%
  group_by(type) %>%
  mutate(true_index = ifelse("True" %in% type, "yes", "no")) %>%
  ggplot(aes(year, N_scaled, group = type)) +
  geom_line(aes(colour = type, lty = true_index)) +
  geom_ribbon(aes(ymin = lwr_scaled, ymax = upr_scaled, fill = type), alpha = 0.3) +
  labs(x = "Year", y = "Relative abundance", colour = "Type", fill = "Type") +
  coord_cartesian(ylim=c(0, 5)) +
  # facet_wrap(~iter, scales = "free_y") +
  theme_minimal() + theme(legend.position = "bottom")

```

# Use part of an actual grid, but make new strata based on depth 
```{r}
plot_grid(survey_grid)
# survey_trim <- crop(survey_grid,
#   extent(400,550,5100,5250))
# 
# values(survey_trim$strat) <- as.numeric(cut(values(survey_trim$depth), seq(0, 1000, by = 100)))

survey_trim <- crop(survey_grid,
  extent(500,700,4953,5150))

values(survey_trim$strat) <- as.numeric(cut(values(survey_trim$depth), seq(0, 1000, by = 150)))
plot_grid(survey_trim) 
```

More variability in depth means a wider depth preference possible now
```{r}
set.seed(42)
 sim <- sim_abundance(ages = seq(1, 10), years = seq(1, 10)) %>%
    sim_distribution(
      grid = survey_trim, 
      ays_covar = sim_ays_covar(
        sd = 2.8, #2.8 is default
        #lambda = 1, # 1 is default, changes degree of smoothness
        range = 150, # make less spatial correlation, from default of 300
        phi_age = 0.8, phi_year = 0.1),
      depth_par = sim_parabola(mu = 200, sigma = 30, plot = F) # simga = 30 was example
    )
```


```{r}
sim$sp_N %>% left_join(sim$grid_xy) %>%
    group_by(x, y, year, cell) %>%
    summarise(N = sum(N), .groups = "drop") %>%
    ggplot(aes(x, y,
      fill = log(N + 1)
      )) +
    geom_raster() +
    facet_wrap(~year) +
    scale_fill_viridis_c()
```

Then we simulate survey samples from this population

```{r}
  survey <- sim_survey(sim, n_sims = 1, min_sets = 1, set_den = 5/1000
    ) %>% run_strat()
  xy <- as_tibble(survey$grid_xy)
  dat <- as_tibble(survey$setdet) %>%
    dplyr::select(x, y, set, year, N = n, tow_area)
  dat <- left_join(dat, xy, by = c("x", "y"))
  dat$offset <- log(dat$tow_area)

  grid_dat <- as_tibble(dplyr::select(sim$grid_xy, x, y, depth)) %>% distinct()
  grid_dat <- purrr::map_dfr(sort(unique(dat$year)), ~ bind_cols(grid_dat, year = .))
  grid_dat$offset <- mean(dat$offset)
```

Here are the survey observations:

```{r}
ggplot(dat, aes(x, y, colour = log(N + 1), size = N)) +
  geom_point() +
  facet_wrap(vars(year)) +
  scale_colour_viridis_c() +
  scale_size_area(max_size = 4)
```



# Fitting a geostatistical spatiotemporal model

```{r}
mesh <- sdmTMB::make_mesh(dat, xy_cols = c("x", "y"), cutoff = 20)
plot(mesh)
```

Fit model with a fixed spatial field that might account for depth even without a true depth covariate
```{r fit2, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
  fit2 <- sdmTMB(N ~ 0 + as.factor(year) + offset,
    data = dat,
    family = nbinom2(link = "log"), spde = mesh,
    include_spatial = TRUE,
    time = "year"
  )
```

```{r}
print(fit2)
```


Fit model with depth, but without a fixed spatial field
```{r fit-depth2, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
 fit_depth2 <- sdmTMB(N ~ 0 + as.factor(year) + s(depth, k = 3) +
      offset,
    data = dat,
    family = nbinom2(link = "log"), spde = mesh,
    # include_spatial = TRUE,
    time = "year"
  )
```

```{r}
print(fit_depth2)
```


```{r}
nd <- data.frame(depth = seq(min(dat$depth), max(dat$depth), length.out = 80))
# add all years in case of time-varying depth
nd <- purrr::map_dfr(sort(unique(dat$year)), ~ bind_cols(nd, year = .))
nd$offset <- mean(dat$offset)
  
  p2 <- predict(fit_depth2,
    newdata = filter(nd, year == nd$year[1]),
    se_fit = TRUE, re_form = NA)
  
 y <- sim_parabola(mu = 200, sigma = 30, plot=F)(x = 50:750)
  tru_curve <- data.frame(x = 50:750, y)  

  ggplot(p2, aes(depth, est,
    ymin = est - 1.96 * est_se, ymax = est + 1.96 * est_se)) +
    geom_line() + geom_ribbon(alpha = 0.4) +
    geom_line(data = tru_curve, aes(y=y, x = x),
      colour = "red", linetype = "dashed", size = 1.5,
      inherit.aes = F) + scale_linetype()
```


And extract values from estimated fixed and random effect variables using the `tidy()` method, which follows the broom package convention:

```{r}
tidy(fit_depth2, conf.int = TRUE)
tidy(fit_depth2, conf.int = TRUE, effects = "ran_pars")
```

We can look at residuals:

```{r}
dat$resid <- residuals(fit_depth2)

ggplot(dat, aes(x, y, colour = resid, size = abs(resid))) +
  geom_point() +
  facet_wrap(~year) +
  scale_colour_gradient2() +
  scale_size_area(max_size = 2)
```

These are randomized quantile residuals---similar to the DHARMa R package. These are adjusted so that they would be Gaussian distributed if the model were consistent with the data:

```{r}
hist(dat$resid)
```

# Predicting over the survey domain and calculating an index of abundance

To calculate an index of abundance, first we need to create a grid that covers the entirety of our survey domain. In this case that would be a square grid:

```{r}
pred <- predict(fit2, newdata = grid_dat, return_tmb_object = TRUE, area = 100)
pred_depth <- predict(fit_depth2, newdata = grid_dat, return_tmb_object = TRUE, area = 100)
index <- get_index(pred)
index_depth <- get_index(pred_depth)

true_abund <- tibble(year = unique(dat$year), N = as.numeric(colSums(survey$I))) %>%
    mutate(type = "True")

strat_abund <- tibble::as_tibble(survey$total_strat) %>%
    mutate(N = total, type = "Design-based") %>%
    dplyr::select(year, N, type)

result <- mutate(index, type = "Model-based", N = est) %>%
    bind_rows(mutate(index_depth, type = "Model-based-depth", N = est)) %>%
    bind_rows(strat_abund) %>%
    bind_rows(true_abund) 

result_scaled <- result %>%
  group_by(type) %>%
  mutate(geo_mean = exp(mean(log(N), na.rm = TRUE)),
    lwr_scaled = lwr / geo_mean, N_scaled = N / geo_mean, upr_scaled = upr / geo_mean,
    type = factor(type, levels = c("True", "Design-based", "Model-based", "Model-based-depth"))) %>%
  ungroup() %>%
  arrange(type)

result_scaled %>%
  group_by(type) %>%
  mutate(true_index = ifelse("True" %in% type, "yes", "no")) %>%
  ggplot(aes(year, N_scaled, group = type)) +
  geom_line(aes(colour = type, lty = true_index)) +
  geom_ribbon(aes(ymin = lwr_scaled, ymax = upr_scaled, fill = type), alpha = 0.3) +
  labs(x = "Year", y = "Relative abundance", colour = "Type", fill = "Type") +
  coord_cartesian(ylim=c(0, 5)) +
  # facet_wrap(~iter, scales = "free_y") +
  theme_minimal() + theme(legend.position = "bottom")

```


```{r}
true <- sim$sp_N %>% left_join(sim$grid_xy) %>%
  group_by(x, y, year, cell) %>% 
  summarise(N = sum(N), .groups = "drop")  %>%
  dplyr::select(x, y, year, N) %>%
  mutate(type = "True")

fitted <- pred_depth$data %>%
  dplyr::select(x, y, year, est) %>%
  mutate(type = "Predicted") %>%
  mutate(N = exp(est))

both <- bind_rows(true, fitted) %>%
  group_by(type) %>%
  mutate(
    N_mean = exp(mean(log(N))),
    N_scaled = N / N_mean
    )

# View(both)

both %>% filter(year >4) %>%
ggplot(aes(x, y, fill = log(N_scaled))) +
  geom_tile() +
  facet_grid(type ~ year) +
  scale_fill_viridis_c(
    # option = "B",
    # end= 0.9,
    # trans = gfranges::fourth_root_power,
    limits = #c(-35, 25),
      quantile(log(both$N_scaled), probs = c(0.025, 1)),
    na.value = "black"
    )
```


```{r}
# is true value within the CI of model-based?
summary_stats1 <- result_scaled %>%
  group_by(year) %>%
  summarise(
    est_lwr = lwr_scaled[type == "Model-based"],
    est_upr = upr_scaled[type == "Model-based"],
    est = N_scaled[type == "Model-based"],
    true = N_scaled[type == "True"], .groups = "drop") %>%
  mutate(covered = est_lwr < true & est_upr > true)
mean(summary_stats1$covered)

# is true value within the CI of model-based with depth?
summary_stats2 <- result_scaled %>%
  group_by(year) %>%
  summarise(
    est_lwr = lwr_scaled[type == "Model-based-depth"],
    est_upr = upr_scaled[type == "Model-based-depth"],
    est = N_scaled[type == "Model-based-depth"],
    true = N_scaled[type == "True"], .groups = "drop") %>%
  mutate(covered = est_lwr < true & est_upr > true)
mean(summary_stats2$covered)

# compare true with predictions as scatter plot
ggplot(summary_stats1, aes(log(true), log(est))) + geom_point() +
  coord_fixed(ylim=c(-1.5, 1.5), xlim = c(-1.5,1.5)) +
  geom_abline(intercept = 0, slope = 1)

ggplot(summary_stats2, aes(log(true), log(est))) + geom_point() +
  coord_fixed(ylim=c(-1.5, 1.5), xlim = c(-1.5,1.5)) +
  geom_abline(intercept = 0, slope = 1)


```



# Fit model with time-varying depth effect
```{r fit-tv, message=FALSE, warning=FALSE, results='hide', cache=TRUE}
 fit_tv_depth <- sdmTMB(N ~ 0 + as.factor(year) + 
      offset,
    data = dat,
    time_varying = ~ 0 + poly(depth, 2),
    family = nbinom2(link = "log"), spde = mesh,
    time = "year"
  )
```


```{r}
p2 <- predict(fit_tv_depth, newdata = nd, se_fit = TRUE, re_form = NA)

ggplot(p2, aes(depth, est,
    ymin = est - 1.96*est_se, ymax = est + 1.96*est_se, colour=as.factor(year))) +
    geom_line() + #geom_ribbon(alpha = 0.4) +
    geom_line(data = tru_curve, aes(y=y, x = x), 
      colour = "red", linetype = "dashed", inherit.aes = F)  +
    scale_colour_viridis_d()
```
